Quantitative Benchmarking for Automatically Generated Conjectures

### 2018-03-14 TEBenchmarkPaper/article.tex
Background cleanup

% For example theory of lists
\newcommand{\function}{\rightarrow}
\newcommand{\Zero}{\text{Z}}
\newcommand{\Succ}{\text{S}}
\newcommand{\List}{\text{List}}
\newcommand{\ListA}{\text{List} \  a}
\newcommand{\Nil}{\text{Nil}}
\newcommand{\Cons}{\text{Cons}}
\newcommand{\Head}{\text{head}}
\newcommand{\Tail}{\text{tail}}
\newcommand{\Append}{\text{append}}
\newcommand{\Reverse}{\text{reverse}}
\newcommand{\Length}{\text{length}}
\newcommand{\Map}{\text{map}}
\newcommand{\Foldl}{\text{foldl}}
\newcommand{\Foldr}{\text{foldr}}

@@

\begin{itemize}
\item A general methodology for benchmarking conjecture generation.
\item Resolving the issue of ``interestingness'' through the use of
  theorem-proving benchmarks as a ground-truth.
\item A specific instantiation of this methodology, using the Tons of Inductive
  Problems benchmark as a corpus.
\item Automated tooling to perform this benchmarking.
\item Application of our methodology to the QuickSpec MTE system, and a
  discussion of the results.
\end{itemize}

Section \ref{sec:background} describes MTE and the conjecture generation problem
in more detail, along with existing approaches and their evaluation
methodologies. We explain our proposal for a more general benchmark in section
\ref{sec:proposal} and section \ref{sec:application} shows the results when
applied to existing MTE tools. Analysis of these results is given in section
\ref{sec:discussion} and concluding remarks in \ref{sec:conclusion}.

@@

Equation \ref{eq:mapreduce} justifies a rewrite rule for splitting a single
$\Map$ call into multiple independent calls dealing with different sections of a
list. Such rules are useful optimisations since each call can be evaluated in
parallel, leading to the ``map/reduce'' programming paradigm.

@@

A major challenge for both ATF and MTE is choosing how to narrow down
the set of generated conjectures to those deemed ``interesting'', since this is
an imprecise term with many different interpretations. For example, all existing
approaches agree that simple tautologies are ``uninteresting'', but differ when
it comes to more complex statements.  Colton \etal{} give a survey of systems for
theory formation and exploration, and their notions of ``interestingness'' for
concepts and conjectures, identifying six notions in
total~\cite{colton2000notion}. These notions, as applied to conjectures, are:

{\bf Empirical plausibility}, which checks whether a property holds
across some specific examples. This is especially useful for avoiding
false conjectures, without resorting to a full proof search.

{\bf Novelty} is whether a conjecture, or one isomorphic or more
general, has already been seen.

@@

\begin{table}
  \centering
  \begin{tabular}{ |l|l|c|c|c|c|c|c| }
    \hline
    \multirow{2}{*}{\textbf{Program}}                     &
    \multirow{2}{*}{\textbf{Conjecture Types}}            &
    \multicolumn{6}{c}{\textbf{Interestingness Measures}} \\ \hhline{~~------}
    \tRow{          &                    & \iE & \iN & \iS & \iA & \iC & \iU}
    \tRow{AM        & \tIFF, \tIMP, \tNE &   X &   X &   X &   X &   X &   X}
    \tRow{GT        & \tIFF, \tIMP, \tNE &   X &   X &   X &   X &   X &   X}
    \tRow{Graffiti  & \tINE              &   X &   X &   X &     &   X &   X}
    \tRow{\Bagai{}  & \tNE               &     &   X &     &   X &   X &    }
    \tRow{HR        & \tIFF, \tIMP, \tNE &   X &   X &   X &   X &   X &   X}
    \tRow{QuickSpec & \tEQ               &   X &   X &     &   X &   X &    }
    \tRow{Speculate & \tCON \tEQ / \tINE &   X &   X &     &   X &   X &    }
    \tRow{IsaCoSy   & \tEQ               &   X &   X &     &   X &   X &   X}
    \tRow{IsaScheme & \tEQ               &   X &   X &     &   X &   X &   X}
  \end{tabular}
  \caption{Classification of MTE systems from~\cite{colton2000notion}, extended
    to include those from~\cite{claessen2013automating} (QuickSpec is the
    conjecture generation component of HipSpec) and the similar Speculate
    system~\cite{braquehais2017speculate}. The interestingness measures are
    \iE{}mpirical plausibility, \iN{}ovelty, \iS{}urprisingness,
    \iA{}pplicability, \iC{}omprehensibility (low complexity) and \iU{}tility.}
  \label{table:colton}
\end{table}

@@

The latter systems all use testing to check for counterexamples, which ensures
results are empirically plausible and applicable (conditions are satisfiable,
types are inhabited, etc.). They also use a small-to-large search order,
ensuring more easily-comprehensibile conjectures are explored first.

IsaCoSy ensures novelty using a constraint system, such that no generated
expression can be simplified using a previously-found equation. The other
systems use post-hoc filters based on unification and term rewriting to discard
expressions or equations which are special-cases of some other, more general
result.

IsaCoSy is able to look for commonly desirable patterns, such as commutativity
and associativity, which increases the utility of its output, whilst IsaScheme
is completely based around such pattern-instantiation.

@@

\subsection{Mathematical Theory Exploration}

@@

Recent systems developed in this area include MATHsAiD~\cite{roy},
IsaScheme~\cite{MontanoRivas2011} and
IsaCoSy~\cite{Johansson.Dixon.Bundy:conjecture-generation}. The goal of the
MATHsAiD (Mechanically Ascertaining Theorems from Hypotheses, Axioms and
Definitions) project is to build a tool which takes in a set of axioms, concept
definitions and a logic and applies its inference rules to reason from the
axioms to theorems. The motivation is to produce a tool which will help
mathematicians to explore the consequences of a set of axioms or a particular
concept definition. One of the main challenges of the project has been to
automatically evaluate interestingness: to distinguish important theorems from
results which, although they follow from a set of axioms, are of little
mathematical interest.

%IsaScheme
Monta{\~n}o-Rivas \etal{} have implemented a scheme-based approach to MTE in their
IsaScheme system~\cite{MontanoRivas2011}. Schemes are higher-order formulae
which can be used to generate new concepts and conjectures; variables within the
scheme are instantiated automatically and this drives the invention process.
For instance, in the theory of natural numbers, given the concepts of successor
($suc$), addition ($+$) and zero ($0$), IsaScheme can use the following scheme
to invent the concept of multiplication:

\[
\text{def-scheme}(g,h,i,j) \equiv\\
  \exists f. \forall x y. \left\{
  \begin{array}{l l}
    f(g,y) = h (y)  \: \text{and}  & \\
    f(i(x),y) = j(y,f(x,y)) & \\
  \end{array} \right.
\]
where the existentially quantified variable $f$ stands for the new
function to be defined in terms of the variables g, h, i and j. Within
the theory of natural numbers, IsaScheme instantiates this scheme with
$\sigma_1 = \{g \mapsto 0, h \mapsto (\lambda x.0), i \mapsto suc, j
\mapsto +\}$. In this example, $f \mapsto *$ (multiplication), since:

\begin{align*}
0*y  &= 0 \\
suc(x) * y &=y + (x*y).
\end{align*}
The new multiplication function $f$ can itself be used to instantiate
variables in the same scheme, resulting in the invention of the
exponentiation concept (these examples are taken from~\cite{MontanoRivas2011}).

%IsaCoSy
The IsaCoSy system (Isabelle Conjecture Synthesis) is a program for
inductive theory formation, which synthesises conjectures from the
available constants and free variables. Only terms which do not
contain subterms that can be rewritten by a set of rewrite rules can
be synthesised.  Conjectures are tested by sending them to a
counterexample checker and, if no counterexamples are found, then sent
to IsaPlanner which attempts to prove them.

% Discussion.

@@

The different backgrounds of the people who named the fields
``automated theory formation'' (Lenat) and ``mathematical theory
exploration'' (Buchberger) perhaps reflects different metaphysical
perspectives on invention (in which case new mathematical ideas are
being {\em formed}, created or produced) and discovery (in which case
abstract mathematical objects exist independently of us and are {\em
  explored} or investigated) in mathematics.\footnote{There is a vast
  literature on the Platonic and the constructivist view of
  mathematics: we shall not venture down this path here (interested
  readers are referred to~\cite{hersh:97,shapiro}). Instead, for now,
  we make the pragmatic assumption that while there may well be
  different cognitive processes involved in invention than those
  involved in discovery, the fields of ATF and MTE are not currently
  concerned with this level of detail and so the philosophical
  distinction is not relevant for our purposes.} As can be seen above,
some of these historical differences are disappearing, and current
systems such as MATHsAiD~\cite{roy}, IsaScheme~\cite{MontanoRivas2011}
and IsaCoSy~\cite{Johansson.Dixon.Bundy:conjecture-generation} are bridging the
methodological gap between ATF and MTE. This is particularly true in terms of
fully-automatic/user-interactive and the domain-specific/general
aspects.

>>>>

% For example theory of lists
\newcommand{\function}{\rightarrow}

\newcommand{\Zero}{\text{Z}}
\newcommand{\Succ}{\text{S}}
\newcommand{\plus}{\text{plus}}
\newcommand{\mult}{\text{times}}

\newcommand{\List}{\text{List}}
\newcommand{\ListA}{\text{List} \  a}
\newcommand{\Nil}{\text{Nil}}
\newcommand{\Cons}{\text{Cons}}
\newcommand{\Head}{\text{head}}
\newcommand{\Tail}{\text{tail}}
\newcommand{\Append}{\text{append}}
\newcommand{\Reverse}{\text{reverse}}
\newcommand{\Length}{\text{length}}
\newcommand{\Map}{\text{map}}
\newcommand{\Foldl}{\text{foldl}}
\newcommand{\Foldr}{\text{foldr}}

@@

\begin{itemize}
\item A general methodology for benchmarking conjecture generation.
\item Resolving the issue of ``interestingness'' through the use of
  theorem-proving benchmarks as a ground-truth.
\item A specific instantiation of this methodology, using the Tons of Inductive
  Problems benchmark as a corpus.
\item Automated tooling to perform this benchmarking.
\item Application of our methodology to the QuickSpec and IsaCoSy MTE systems,
  and a comparison and discussion of the results.
\end{itemize}

Section \ref{sec:background} describes MTE and the conjecture generation problem
in more detail, along with existing approaches and their evaluation
methodologies. We explain our proposal for a more general benchmark in section
\ref{sec:proposal} whilst section \ref{sec:application} shows the results when
applied to existing MTE tools. Analysis of these results is given in section
\ref{sec:discussion} and concluding remarks in section \ref{sec:conclusion}.

@@

Equation \ref{eq:mapreduce} justifies a rewrite rule for splitting a single
$\Map$ call into multiple independent calls dealing with different sections of a
list. Such rules can be useful optimisations, for example in a distributed
context where since each call can be evaluated in parallel (this is the basis of
the ``map/reduce'' programming paradigm).

@@

A major challenge for both ATF and MTE is choosing how to narrow down
the set of generated conjectures to those deemed ``interesting'', since this is
an imprecise term with many different interpretations. For example, all existing
approaches agree that simple tautologies are ``uninteresting'', but differ when
it comes to more complex statements.  Colton \etal{} give a survey of systems for
theory formation and exploration, and their associated notions of
``interestingness'' for concepts and conjectures~\cite{colton2000notion}. Six
notions are identified, which are applied to conjectures as follows:

{\bf Empirical plausibility} checks whether a property holds across some
specific examples. This is especially useful for avoiding false conjectures,
without resorting to a full proof search.

{\bf Novelty} depends on whether a conjecture, or one isomorphic or more
general, has already been seen.

@@

\begin{table}
  \centering
  \begin{tabular}{ |l|l|c|c|c|c|c|c| }
    \hline
    \multirow{2}{*}{\textbf{Program}}                     &
    \multirow{2}{*}{\textbf{Conjecture Types}}            &
    \multicolumn{6}{c}{\textbf{Interestingness Measures}} \\ \hhline{~~------}
    \tRow{          &                    & \iE & \iN & \iS & \iA & \iC & \iU}
    \tRow{AM        & \tIFF, \tIMP, \tNE &   X &   X &   X &   X &   X &   X}
    \tRow{GT        & \tIFF, \tIMP, \tNE &   X &   X &   X &   X &   X &   X}
    \tRow{Graffiti  & \tINE              &   X &   X &   X &     &   X &   X}
    \tRow{\Bagai{}  & \tNE               &     &   X &     &   X &   X &    }
    \tRow{HR        & \tIFF, \tIMP, \tNE &   X &   X &   X &   X &   X &   X}
    \tRow{QuickSpec & \tEQ               &   X &   X &     &   X &   X &    }
    \tRow{Speculate & \tCON \tEQ / \tINE &   X &   X &     &   X &   X &    }
    \tRow{IsaCoSy   & \tEQ               &   X &   X &     &   X &   X &   X}
    \tRow{IsaScheme & \tEQ               &   X &   X &     &   X &   X &   X}
  \end{tabular}
  \caption{Classification of MTE systems from~\cite{colton2000notion}, extended
    to include those from~\cite{claessen2013automating} (QuickSpec is the
    conjecture generation component of HipSpec) and the similar Speculate
    system~\cite{braquehais2017speculate}. The interestingness measures are
    to include four more recent systems. The interestingness measures are
    \iE{}mpirical plausibility, \iN{}ovelty, \iS{}urprisingness,
    \iA{}pplicability, \iC{}omprehensibility (low complexity) and \iU{}tility.}
  \label{table:colton}
\end{table}

@@

The latter systems are described in more detail in section~\ref{section:mte},
but all use testing to check for counterexamples, which ensures results are
empirically plausible and applicable (conditions are satisfiable, types are
inhabited, etc.). They also use a small-to-large search order, ensuring more
easily-comprehensibile conjectures are explored first.

IsaCoSy ensures novelty using a constraint system to prevent special-cases being
generated, whilst the others use post-hoc filters based on unification and term
rewriting to achieve the same result. IsaCoSy is able to look for commonly
desirable patterns, such as commutativity and associativity, which increases the
utility of its output, whilst IsaScheme is completely based around such
pattern-instantiation.

@@

\subsection{Mathematical Theory Exploration}
\label{section:mte}

@@

Recent systems developed in this area include IsaScheme~\cite{MontanoRivas2011},
IsaCoSy~\cite{Johansson.Dixon.Bundy:conjecture-generation},
QuickSpec~\cite{QuickSpec} and Speculate~\cite{braquehais2017speculate}
(included in table~\ref{table:colton}) as well as MATHsAiD~\cite{roy}.

The goal of the MATHsAiD (Mechanically Ascertaining Theorems from Hypotheses,
Axioms and Definitions) project is to build a tool which takes in a set of
axioms, concept definitions and a logic and applies its inference rules to
reason from the axioms to theorems. The motivation is to produce a tool which
will help mathematicians to explore the consequences of a set of axioms or a
particular concept definition. One of the main challenges of the project has
been to automatically evaluate interestingness: to distinguish important
theorems from results which, although they follow from a set of axioms, are of
little mathematical interest.

%IsaScheme
Monta{\~n}o-Rivas \etal{} have implemented a scheme-based approach to MTE in their
IsaScheme system~\cite{MontanoRivas2011}. Schemes are higher-order formulae
which can be used to generate new concepts and conjectures; variables within the
scheme are instantiated automatically and this drives the invention process.
For instance, in the theory of natural numbers, given the concepts of zero
($\Zero$), successor ($\Succ$) and addition ($\plus$), IsaScheme can use the
following scheme to invent the concept of multiplication:

\[
\text{def-scheme}(g,h,i,j) \equiv\\
  \exists f. \forall x y. \left\{
  \begin{array}{l l}
    f(g,y) = h (y)  \: \text{and}  & \\
    f(i(x),y) = j(y,f(x,y)) & \\
  \end{array} \right.
\]

where the existentially quantified variable $f$ stands for the new
function to be defined in terms of the variables $g$, $h$, $i$ and $j$. Within
the theory of natural numbers, IsaScheme instantiates this scheme with
$\sigma_1 = \{g \mapsto \Zero, h \mapsto (\lambda x.\Zero), i \mapsto \Succ, j
\mapsto \plus\}$. In this example, $f \mapsto \mult$ (multiplication), since:

\begin{align*}
\mult(\Zero, y)  &= \Zero \\
\mult(\Succ(x), y) &= \plus(y, \mult(x, y))
\end{align*}

The new multiplication function $f$ can itself be used to instantiate
variables in the same scheme, resulting in the invention of the
exponentiation concept (these examples are taken from~\cite{MontanoRivas2011}).

%IsaCoSy
The IsaCoSy system (Isabelle Conjecture Synthesis) conjectures equations by
enumerating expressions involving a given set of (typed) constants and free
variables (a signature). A constraint solver forbids certain (sub)expressions
from being synthesised, and these constraints are extended whenever a new
conjecture is generated, to avoid generating any special-cases of this
conjecture in the future. Conjectures are tested by sending them to the
QuickCheck counterexample finder and, if no counterexamples are found, then
sent to IsaPlanner which attempts to prove them.

%QuickSpec
QuickSpec emerged from work on the QuickCheck software testing framework for the
Haskell programming language, and has been used to generate conjectures for the
subsequent HipSpec and Hipster MTE systems. Like IsaCoSy, a signature is
provided and terms are enumerated. Rather than a constraint system, QuickSpec
version 1 collects expressions of the same type into equivalence classes,
assuming them to be equal. QuickCheck is used to test this assumption by
instantiating each variable with a randomly chosen value, evaluating the
resulting expressions and comparing them. Any equivalence class whose elements
don't compare equal are split up to separate those observably-different
expressions, and the process is repeated with new random values.

After 500 rounds of testing, any expressions still sharing the same equivalence
class are conjectured to be equal for all values of their variables. Finally, a
congruence closure algorithm is applied to these equations, to remove
redundancies and special-cases.

QuickSpec version 2 replaces this two-step process with a single, iterative
algorithm similar to that of IsaCoSy. Generated conjectures are fed into a
Knuth-Bendix completion algorithm to form a corresponding set of rewrite rules.
As expressions are enumerated, they are simplified using these rules and
discarded if equal to a known expression. If not, QuickCheck tests whether the
new expression can be distinguished from the known expressions through random
testing: those which can are added to the set of known expressions. Those which
cannot be distinguished are conjectured to be equal, and the rewrite rules are
updated.

%Speculate
Speculate is a Haskell tool which operates in a similar way to QuickSpec
version 2. In addition to equations, Speculate uses the laws of total orders and
Boolean algebra, along with testing (via LeanCheck), to conjecture inequalities
and conditional relations between expressions, such as (for
$x, y \in \mathbb{Z}$):

\begin{equation*}
  \begin{aligned}
    x + y \leq \lvert x \rvert + \lvert y \rvert \\
    x \leq 0 \rightarrow x + \lvert x \rvert = 0
  \end{aligned}
\end{equation*}

@@

The different backgrounds of the people who named the fields
``automated theory formation'' (Lenat) and ``mathematical theory
exploration'' (Buchberger) perhaps reflects different metaphysical
perspectives on invention (in which case new mathematical ideas are
being {\em formed}, created or produced) and discovery (in which case
abstract mathematical objects exist independently of us and are {\em
  explored} or investigated) in mathematics.\footnote{There is a vast
  literature on the Platonic and the constructivist view of
  mathematics: we shall not venture down this path here (interested
  readers are referred to~\cite{hersh:97,shapiro}). Instead, for now,
  we make the pragmatic assumption that while there may well be
  different cognitive processes involved in invention than those
  involved in discovery, the fields of ATF and MTE are not currently
  concerned with this level of detail and so the philosophical
  distinction is not relevant for our purposes.} As can be seen above,
some of these historical differences are disappearing, and recent
implementations are bridging the methodological gap between ATF and MTE. This is
particularly true in terms of fully-automatic/user-interactive and the
domain-specific/general aspects.
